---
title: "MGEO RGC/IPC Monocle3 Interactive Notebook"
output: html_notebook
---

Use this on a Great Lakes compute node after starting the interactive allocation
with `python_notebooks/scripts/start_vscode_r_monocle_compute_job.sh`.

The VS Code workspace is configured to start R through
`scripts/r_monocle3_vscode.sh`, which loads:

```bash
module load Bioinformatics
module load Rmonocle3/1.3.7
```

```{r setup}
options(stringsAsFactors = FALSE)
options(expressions = 500000)

suppressPackageStartupMessages({
  library(Matrix)
  library(monocle3)
})

PROJECT_ROOT <- Sys.getenv(
  "PROJECT_ROOT",
  unset = "/nfs/turbo/umms-parent/mgeo_neuron_scrnaseq_projectfolder"
)

RUN_ROOT <- file.path(PROJECT_ROOT, "results/mgeo_rgc_ipc_monocle3")
INPUT_DIR <- file.path(RUN_ROOT, "inputs")
INTERACTIVE_DIR <- file.path(RUN_ROOT, "interactive")
PLOT_DIR <- file.path(INTERACTIVE_DIR, "plots")
dir.create(INTERACTIVE_DIR, recursive = TRUE, showWarnings = FALSE)
dir.create(PLOT_DIR, recursive = TRUE, showWarnings = FALSE)

EXPR_MTX_PATH <- file.path(INPUT_DIR, "mgeo_rgc_ipc_expression_genes_by_cells.mtx")
CELL_METADATA_PATH <- file.path(INPUT_DIR, "mgeo_rgc_ipc_cell_metadata.csv")
GENE_METADATA_PATH <- file.path(INPUT_DIR, "mgeo_rgc_ipc_gene_metadata.csv")

CDS_PREPROCESS_RDS <- file.path(INTERACTIVE_DIR, "cds_after_preprocess.rds")
CDS_UMAP_CLUSTER_RDS <- file.path(INTERACTIVE_DIR, "cds_after_umap_cluster_no_graph.rds")
UMAP_CLUSTER_CSV <- file.path(INTERACTIVE_DIR, "mgeo_rgc_ipc_monocle3_umap_clusters.csv")
CDS_UMAP_RDS <- file.path(INTERACTIVE_DIR, "cds_after_umap_cluster_graph.rds")
PSEUDOTIME_CSV <- file.path(INTERACTIVE_DIR, "mgeo_rgc_ipc_monocle3_pseudotime_interactive.csv")
CDS_ORDERED_RDS <- file.path(INTERACTIVE_DIR, "cds_ordered_interactive.rds")

NUM_DIM <- 50
LEARN_GRAPH_USE_PARTITION <- TRUE

log_step <- function(...) {
  message(format(Sys.time(), "%Y-%m-%d %H:%M:%S"), " | ", paste0(..., collapse = ""))
  flush.console()
}

timed_step <- function(label, expr) {
  log_step("START ", label)
  start_time <- proc.time()[["elapsed"]]
  result <- force(expr)
  elapsed_min <- (proc.time()[["elapsed"]] - start_time) / 60
  log_step("DONE ", label, " (", round(elapsed_min, 2), " min)")
  result
}

safe_save_rds <- function(object, path) {
  tryCatch(
    timed_step(paste0("save RDS ", basename(path)), {
      saveRDS(object, path, compress = FALSE)
    }),
    error = function(e) {
      warning(
        "Could not save RDS checkpoint to ",
        path,
        ": ",
        conditionMessage(e),
        immediate. = TRUE
      )
      invisible(FALSE)
    }
  )
}

get_umap_clusters <- function(cds) {
  monocle3::clusters(cds, reduction_method = "UMAP")
}

get_umap_partitions <- function(cds) {
  monocle3::partitions(cds, reduction_method = "UMAP")
}

write_umap_cluster_checkpoint <- function(cds, path) {
  umap <- reducedDims(cds)$UMAP
  cluster_values <- get_umap_clusters(cds)
  checkpoint <- data.frame(
    cell_id = rownames(umap),
    monocle3_umap_1 = umap[, 1],
    monocle3_umap_2 = umap[, 2],
    monocle3_cluster = as.character(cluster_values),
    stringsAsFactors = FALSE
  )
  timed_step(paste0("write CSV ", basename(path)), {
    write.csv(checkpoint, path, row.names = FALSE)
  })
}

has_umap <- function(cds) {
  "UMAP" %in% names(reducedDims(cds)) &&
    !is.null(reducedDims(cds)$UMAP) &&
    nrow(reducedDims(cds)$UMAP) == ncol(cds)
}

has_umap_clusters <- function(cds) {
  cluster_values <- tryCatch(
    get_umap_clusters(cds),
    error = function(e) NULL
  )
  !is.null(cluster_values) && length(cluster_values) == ncol(cds)
}

PROJECT_ROOT
INPUT_DIR
INTERACTIVE_DIR
```

## Runtime Diagnostics

Run this chunk before doing any heavy work. It confirms that the notebook is
running on the allocated compute node, using the Monocle3 module R, and seeing
the expected project paths.

```{r runtime-diagnostics}
cat("=== Host / Slurm ===\n")
cat("hostname:", Sys.info()[["nodename"]], "\n")
cat("user:", Sys.info()[["user"]], "\n")
cat("SLURM_JOB_ID:", Sys.getenv("SLURM_JOB_ID", unset = "<unset>"), "\n")
cat("SLURM_JOB_NAME:", Sys.getenv("SLURM_JOB_NAME", unset = "<unset>"), "\n")
cat("SLURM_CPUS_PER_TASK:", Sys.getenv("SLURM_CPUS_PER_TASK", unset = "<unset>"), "\n")
cat("SLURM_MEM_PER_NODE:", Sys.getenv("SLURM_MEM_PER_NODE", unset = "<unset>"), "\n")
cat("SLURM_JOB_NODELIST:", Sys.getenv("SLURM_JOB_NODELIST", unset = "<unset>"), "\n")

cat("\n=== R ===\n")
cat("R.home():", R.home(), "\n")
cat("R.version:", R.version.string, "\n")
cat("R executable from shell:", system("which R", intern = TRUE), "\n")
cat("Rscript executable from shell:", system("which Rscript", intern = TRUE), "\n")
cat(".libPaths():\n")
print(.libPaths())

cat("\n=== Modules ===\n")
module_list <- tryCatch(
  system("module list 2>&1", intern = TRUE),
  error = function(e) paste("module list failed:", conditionMessage(e))
)
cat(paste(module_list, collapse = "\n"), "\n")

cat("\n=== Package Availability ===\n")
pkgs <- c("monocle3", "Matrix", "igraph", "HDF5Array", "sf", "spdep")
print(data.frame(package = pkgs, installed = sapply(pkgs, requireNamespace, quietly = TRUE)), row.names = FALSE)

cat("\n=== Parallel / Memory Hints ===\n")
cat("parallel::detectCores():", parallel::detectCores(), "\n")
meminfo <- tryCatch(readLines("/proc/meminfo", n = 3), error = function(e) character(0))
cat(paste(meminfo, collapse = "\n"), "\n")

cat("\n=== Paths ===\n")
cat("getwd():", getwd(), "\n")
cat("PROJECT_ROOT:", PROJECT_ROOT, "\n")
cat("INPUT_DIR:", INPUT_DIR, "\n")
cat("INTERACTIVE_DIR:", INTERACTIVE_DIR, "\n")
cat("Expression matrix exists:", file.exists(EXPR_MTX_PATH), "\n")
cat("Cell metadata exists:", file.exists(CELL_METADATA_PATH), "\n")
cat("Gene metadata exists:", file.exists(GENE_METADATA_PATH), "\n")
if (file.exists(EXPR_MTX_PATH)) {
  cat("Expression matrix GB:", round(file.info(EXPR_MTX_PATH)$size / 1e9, 3), "\n")
}

stopifnot(requireNamespace("monocle3", quietly = TRUE))
stopifnot(grepl("^gl[0-9]+", Sys.info()[["nodename"]]))

if (Sys.getenv("SLURM_JOB_ID", unset = "") == "") {
  warning(
    "SLURM_JOB_ID is unset. This can happen when VS Code launches a fresh ",
    "R Markdown render process on a compute node instead of using the R ",
    "terminal that inherited salloc. Hostname is still a compute node: ",
    Sys.info()[["nodename"]]
  )
}
```

## Read Inputs

```{r read-inputs}
stopifnot(file.exists(EXPR_MTX_PATH))
stopifnot(file.exists(CELL_METADATA_PATH))
stopifnot(file.exists(GENE_METADATA_PATH))

log_step("Reading expression matrix from ", EXPR_MTX_PATH)
expression_matrix <- timed_step("Matrix::readMM", {
  Matrix::readMM(EXPR_MTX_PATH)
})
if (!inherits(expression_matrix, "dgCMatrix")) {
  expression_matrix <- timed_step("Convert expression matrix to dgCMatrix", {
    as(expression_matrix, "dgCMatrix")
  })
}

log_step("Reading cell metadata from ", CELL_METADATA_PATH)
cell_metadata <- timed_step("read cell metadata CSV", {
  read.csv(CELL_METADATA_PATH, row.names = 1, check.names = FALSE)
})

log_step("Reading gene metadata from ", GENE_METADATA_PATH)
gene_metadata <- timed_step("read gene metadata CSV", {
  read.csv(GENE_METADATA_PATH, row.names = 1, check.names = FALSE)
})

stopifnot(nrow(expression_matrix) == nrow(gene_metadata))
stopifnot(ncol(expression_matrix) == nrow(cell_metadata))

rownames(expression_matrix) <- rownames(gene_metadata)
colnames(expression_matrix) <- rownames(cell_metadata)

log_step(
  "Loaded matrix with ", nrow(expression_matrix), " genes x ",
  ncol(expression_matrix), " cells; nonzero entries = ", length(expression_matrix@x)
)

dim(expression_matrix)
head(cell_metadata[, c("DIV", "shi_s5_RGC_score", "shi_s5_IPC_score")])
```

## Create CDS And Preprocess

```{r preprocess}
cds <- timed_step("new_cell_data_set", {
  new_cell_data_set(
    expression_matrix,
    cell_metadata = cell_metadata,
    gene_metadata = gene_metadata
  )
})

set.seed(7)
cds <- timed_step(paste0("preprocess_cds num_dim=", NUM_DIM), {
  preprocess_cds(cds, num_dim = NUM_DIM, norm_method = "none")
})
safe_save_rds(cds, CDS_PREPROCESS_RDS)
CDS_PREPROCESS_RDS
```

## Resume From Preprocess Checkpoint

Use this chunk after restarting R or moving to a new compute node. It lets you
skip rereading the matrix and rerunning `preprocess_cds()` if
`cds_after_preprocess.rds` exists.

```{r resume-preprocess}
stopifnot(file.exists(CDS_PREPROCESS_RDS))

cds <- timed_step("read preprocessed CDS RDS", {
  readRDS(CDS_PREPROCESS_RDS)
})

cell_metadata <- as.data.frame(colData(cds))
gene_metadata <- as.data.frame(rowData(cds))

log_step("Loaded preprocessed CDS: ", nrow(cds), " genes x ", ncol(cds), " cells")
dim(cds)
```

## UMAP, Clusters, Graph

```{r graph}
set.seed(7)

if (!has_umap(cds)) {
  cds <- timed_step("reduce_dimension UMAP", {
    reduce_dimension(cds, reduction_method = "UMAP")
  })
} else {
  log_step("Skipping reduce_dimension; UMAP already exists.")
}
log_step("UMAP coordinates: ", paste(dim(reducedDims(cds)$UMAP), collapse = " x "))

if (!has_umap_clusters(cds)) {
  cds <- timed_step("cluster_cells UMAP", {
    cluster_cells(cds, reduction_method = "UMAP")
  })
} else {
  log_step("Skipping cluster_cells; UMAP clusters already exist.")
}
log_step("Cluster count: ", length(unique(get_umap_clusters(cds))))

write_umap_cluster_checkpoint(cds, UMAP_CLUSTER_CSV)
safe_save_rds(cds, CDS_UMAP_CLUSTER_RDS)

cds <- tryCatch(
  timed_step(paste0("learn_graph use_partition=", LEARN_GRAPH_USE_PARTITION), {
    learn_graph(cds, use_partition = LEARN_GRAPH_USE_PARTITION)
  }),
  error = function(e) {
    write_umap_cluster_checkpoint(cds, UMAP_CLUSTER_CSV)
    safe_save_rds(cds, CDS_UMAP_CLUSTER_RDS)
    stop(
      "learn_graph failed after saving the UMAP/cluster checkpoint to ",
      UMAP_CLUSTER_CSV,
      ". Error: ",
      conditionMessage(e),
      call. = FALSE
    )
  }
)
log_step("Principal graph nodes: ", length(igraph::V(principal_graph(cds)[["UMAP"]])))
log_step("Principal graph edges: ", length(igraph::E(principal_graph(cds)[["UMAP"]])))

safe_save_rds(cds, CDS_UMAP_RDS)
CDS_UMAP_RDS
```

```{r plot-div, fig.width=8, fig.height=6}
plot_cells(
  cds,
  color_cells_by = "DIV",
  label_groups_by_cluster = FALSE,
  label_leaves = FALSE,
  label_branch_points = FALSE
)
```

```{r plot-rgc, fig.width=8, fig.height=6}
plot_cells(
  cds,
  color_cells_by = "shi_s5_RGC_score",
  label_groups_by_cluster = FALSE,
  label_leaves = FALSE,
  label_branch_points = FALSE
)
```

```{r plot-ipc, fig.width=8, fig.height=6}
plot_cells(
  cds,
  color_cells_by = "shi_s5_IPC_score",
  label_groups_by_cluster = FALSE,
  label_leaves = FALSE,
  label_branch_points = FALSE
)
```

## Root Candidates

```{r root-candidates}
root_cells <- rownames(cell_metadata)[as.logical(cell_metadata$monocle_root_candidate)]
root_seed_cell <- rownames(cell_metadata)[as.logical(cell_metadata$monocle_root_seed_cell)]

length(root_cells)
root_seed_cell

root_summary <- cell_metadata[root_cells, c(
  "DIV",
  "shi_s5_RGC_score",
  "shi_s5_IPC_score",
  "shi_s5_IPC_minus_RGC_score"
)]
summary(root_summary)
table(root_summary$DIV)
```

## Root Principal Nodes

```{r root-nodes}
get_root_pr_nodes <- function(cds, root_cell_ids) {
  closest_vertex <- cds@principal_graph_aux[["UMAP"]]$pr_graph_cell_proj_closest_vertex
  closest_vertex <- as.matrix(closest_vertex[colnames(cds), , drop = FALSE])
  available_root_cells <- intersect(root_cell_ids, rownames(closest_vertex))
  if (length(available_root_cells) == 0) {
    stop("No root cells represented in principal graph mapping.")
  }
  root_vertex_table <- sort(table(closest_vertex[available_root_cells, 1]), decreasing = TRUE)
  root_vertex_ids <- as.numeric(names(root_vertex_table))
  graph_node_names <- igraph::V(principal_graph(cds)[["UMAP"]])$name[root_vertex_ids]
  list(
    table = root_vertex_table,
    primary = graph_node_names[[1]],
    all = graph_node_names
  )
}

root_nodes <- get_root_pr_nodes(cds, root_cells)
root_nodes$table[1:20]
root_nodes$primary
```

## Order Cells

```{r order-cells}
cds_ordered <- tryCatch(
  order_cells(cds, reduction_method = "UMAP", root_cells = root_cells),
  error = function(e) {
    message("order_cells(root_cells=...) failed: ", conditionMessage(e))
    NULL
  }
)

if (is.null(cds_ordered)) {
  cds_ordered <- order_cells(
    cds,
    reduction_method = "UMAP",
    root_pr_nodes = root_nodes$primary
  )
}

pt <- monocle3::pseudotime(cds_ordered)
summary(pt)
table(is.finite(pt), useNA = "ifany")
```

```{r plot-pseudotime, fig.width=8, fig.height=6}
plot_cells(
  cds_ordered,
  color_cells_by = "pseudotime",
  label_groups_by_cluster = FALSE,
  label_leaves = TRUE,
  label_branch_points = TRUE
)
```

## Export For Python

```{r export}
pt_df <- data.frame(
  cell_id = names(pt),
  rgc_ipc_pseudotime = as.numeric(pt),
  rgc_ipc_pseudotime_method = paste0(
    "monocle3_interactive_use_partition_",
    LEARN_GRAPH_USE_PARTITION
  ),
  monocle3_partition = as.character(get_umap_partitions(cds_ordered)),
  monocle3_cluster = as.character(get_umap_clusters(cds_ordered)),
  monocle3_root_pr_node = root_nodes$primary,
  stringsAsFactors = FALSE
)

keep_meta <- intersect(
  c(
    "DIV",
    "original_cell_id",
    "monocle_root_candidate",
    "monocle_root_seed_cell",
    "shi_s5_RGC_score",
    "shi_s5_IPC_score",
    "shi_s5_IPC_minus_RGC_score"
  ),
  colnames(cell_metadata)
)
pt_df <- cbind(pt_df, cell_metadata[pt_df$cell_id, keep_meta, drop = FALSE])

write.csv(pt_df, PSEUDOTIME_CSV, row.names = FALSE)
safe_save_rds(cds_ordered, CDS_ORDERED_RDS)

PSEUDOTIME_CSV
CDS_ORDERED_RDS
```

## Marker Validation Figures

This section uses the existing ordered object only. It does not recompute UMAP,
graph learning, roots, or pseudotime.

```{r marker-validation-figures, fig.width=12, fig.height=10}
stopifnot(exists("cds_ordered"))
stopifnot("UMAP" %in% names(reducedDims(cds_ordered)))

suppressPackageStartupMessages({
  library(ggplot2)
})

marker_genes <- c(
  "NES",
  "VIM",
  "HES1",
  "ASCL1",
  "DLX2",
  "DCX",
  "STMN2",
  "FABP7",
  "HES5",
  "SOX2",
  "DLX1",
  "NKX2-1"
)

available_marker_genes <- intersect(marker_genes, rownames(cds_ordered))
missing_marker_genes <- setdiff(marker_genes, available_marker_genes)
if (length(missing_marker_genes) > 0) {
  warning("Missing marker genes: ", paste(missing_marker_genes, collapse = ", "))
}
stopifnot(length(available_marker_genes) > 0)

umap <- reducedDims(cds_ordered)$UMAP
umap_df <- data.frame(
  cell_id = rownames(umap),
  UMAP_1 = umap[, 1],
  UMAP_2 = umap[, 2],
  pseudotime = as.numeric(monocle3::pseudotime(cds_ordered)),
  stringsAsFactors = FALSE
)

expr_mat <- exprs(cds_ordered)[available_marker_genes, umap_df$cell_id, drop = FALSE]
if (inherits(expr_mat, "sparseMatrix")) {
  expr_mat <- as.matrix(expr_mat)
}

scale_to_unit <- function(x) {
  finite_x <- x[is.finite(x)]
  if (length(finite_x) == 0) {
    return(rep(NA_real_, length(x)))
  }
  lo <- min(finite_x, na.rm = TRUE)
  hi <- stats::quantile(finite_x, 0.99, na.rm = TRUE, names = FALSE)
  if (!is.finite(hi) || hi <= lo) {
    return(ifelse(is.finite(x), 0, NA_real_))
  }
  pmin(pmax((x - lo) / (hi - lo), 0), 1)
}

scale_to_limit <- function(x, upper) {
  if (!is.finite(upper) || upper <= 0) {
    return(rep(0, length(x)))
  }
  pmin(pmax(x / upper, 0), 1)
}

get_principal_graph_edges <- function(cds) {
  graph <- principal_graph(cds)[["UMAP"]]
  coords <- cds@principal_graph_aux[["UMAP"]]$dp_mst
  if (is.null(graph) || is.null(coords)) {
    return(data.frame())
  }
  if (nrow(coords) != 2 && ncol(coords) == 2) {
    coords <- t(coords)
  }
  edges <- igraph::as_data_frame(graph, what = "edges")
  node_names <- igraph::V(graph)$name
  if (is.null(colnames(coords))) {
    colnames(coords) <- node_names
  }
  edges <- edges[edges$from %in% colnames(coords) & edges$to %in% colnames(coords), ]
  data.frame(
    x = coords[1, edges$from],
    y = coords[2, edges$from],
    xend = coords[1, edges$to],
    yend = coords[2, edges$to],
    stringsAsFactors = FALSE
  )
}

graph_edges <- get_principal_graph_edges(cds_ordered)

expression_cap <- function(x, q = 0.99) {
  finite_x <- pmax(x[is.finite(x)], 0)
  if (length(finite_x) == 0) return(1)
  cap <- stats::quantile(finite_x, q, na.rm = TRUE, names = FALSE)
  if (!is.finite(cap) || cap <= 0) {
    cap <- max(finite_x, na.rm = TRUE)
  }
  max(ceiling(cap), 1)
}

developmental_colors <- c("grey92", "gold", "orange", "firebrick")
base_umap_theme <- function() {
  theme_void(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 13, hjust = 0.5),
      legend.position = "right",
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8),
      plot.margin = margin(3, 3, 3, 3)
    )
}

pseudotime_cap <- max(ceiling(stats::quantile(
  umap_df$pseudotime[is.finite(umap_df$pseudotime)],
  0.99,
  na.rm = TRUE,
  names = FALSE
)), 1)
pseudotime_plot_df <- transform(
  umap_df,
  plot_value = pmin(pmax(pseudotime, 0), pseudotime_cap),
  alpha_value = ifelse(is.finite(pseudotime), 0.85, 0.03)
)
pseudotime_umap_plot <- ggplot(pseudotime_plot_df, aes(UMAP_1, UMAP_2)) +
  geom_point(aes(color = plot_value, alpha = alpha_value), size = 0.09, stroke = 0) +
  geom_segment(
    data = graph_edges,
    aes(x = x, y = y, xend = xend, yend = yend),
    inherit.aes = FALSE,
    color = "grey30",
    linewidth = 0.11,
    alpha = 0.38
  ) +
  scale_color_gradientn(
    colors = developmental_colors,
    limits = c(0, pseudotime_cap),
    breaks = seq(0, pseudotime_cap, by = max(1, ceiling(pseudotime_cap / 4))),
    na.value = "grey92",
    name = "pseudotime"
  ) +
  scale_alpha(range = c(0.025, 0.95), limits = c(0, 1), guide = "none") +
  coord_equal() +
  labs(title = "pseudotime", x = NULL, y = NULL) +
  base_umap_theme()

gene_umap_plots <- list()
gene_umap_df <- data.frame()
for (gene in available_marker_genes) {
  expr_values <- as.numeric(expr_mat[gene, ])
  gene_cap <- expression_cap(expr_values, q = 0.99)
  gene_df <- data.frame(
    cell_id = umap_df$cell_id,
    UMAP_1 = umap_df$UMAP_1,
    UMAP_2 = umap_df$UMAP_2,
    panel = gene,
    value = expr_values,
    plot_value = pmin(pmax(expr_values, 0), gene_cap),
    alpha_value = scale_to_limit(expr_values, gene_cap),
    stringsAsFactors = FALSE
  )
  gene_umap_df <- rbind(gene_umap_df, gene_df)
  gene_umap_plots[[gene]] <- ggplot(gene_df, aes(UMAP_1, UMAP_2)) +
    geom_point(aes(color = plot_value, alpha = alpha_value), size = 0.09, stroke = 0) +
    geom_segment(
      data = graph_edges,
      aes(x = x, y = y, xend = xend, yend = yend),
      inherit.aes = FALSE,
      color = "grey30",
      linewidth = 0.11,
      alpha = 0.38
    ) +
    scale_color_gradientn(
      colors = developmental_colors,
      limits = c(0, gene_cap),
      breaks = seq(0, gene_cap, by = 1),
      na.value = "grey92",
      name = "log expr"
    ) +
    scale_alpha(range = c(0.025, 0.95), limits = c(0, 1), guide = "none") +
    coord_equal() +
    labs(title = gene, x = NULL, y = NULL) +
    base_umap_theme()
}

umap_grid <- gridExtra::arrangeGrob(
  grobs = c(list(pseudotime_umap_plot), gene_umap_plots),
  ncol = 3
)
gene_umap_grid <- gridExtra::arrangeGrob(grobs = gene_umap_plots, ncol = 3)

gene_clip_summary <- data.frame(
  gene = available_marker_genes,
  clip_quantile = 0.99,
  clip_limit_log_expr = vapply(
    available_marker_genes,
    function(gene) expression_cap(as.numeric(expr_mat[gene, ]), q = 0.99),
    numeric(1)
  ),
  stringsAsFactors = FALSE
)
gene_clip_summary_path <- file.path(PLOT_DIR, "monocle3_marker_gene_color_limits.csv")
write.csv(gene_clip_summary, gene_clip_summary_path, row.names = FALSE)

umap_png <- file.path(PLOT_DIR, "monocle3_marker_umap_grid.png")
umap_pdf <- file.path(PLOT_DIR, "monocle3_marker_umap_grid.pdf")
ggsave(umap_png, umap_grid, width = 15, height = 18, dpi = 240)
ggsave(umap_pdf, umap_grid, width = 15, height = 18, device = cairo_pdf)

gene_umap_png <- file.path(PLOT_DIR, "monocle3_marker_gene_umap_grid.png")
gene_umap_pdf <- file.path(PLOT_DIR, "monocle3_marker_gene_umap_grid.pdf")
ggsave(gene_umap_png, gene_umap_grid, width = 15, height = 16, dpi = 240)
ggsave(gene_umap_pdf, gene_umap_grid, width = 15, height = 16, device = cairo_pdf)

finite_pt <- is.finite(umap_df$pseudotime)
pseudotime_gene_df <- data.frame()
for (gene in available_marker_genes) {
  pseudotime_gene_df <- rbind(
    pseudotime_gene_df,
    data.frame(
      cell_id = umap_df$cell_id[finite_pt],
      pseudotime = umap_df$pseudotime[finite_pt],
      gene = gene,
      expression = as.numeric(expr_mat[gene, finite_pt]),
      stringsAsFactors = FALSE
    )
  )
}
pseudotime_gene_df$gene <- factor(
  pseudotime_gene_df$gene,
  levels = marker_genes[marker_genes %in% available_marker_genes]
)

pseudotime_grid <- ggplot(pseudotime_gene_df, aes(pseudotime, expression)) +
  geom_point(color = "grey75", alpha = 0.045, size = 0.08, stroke = 0) +
  geom_smooth(
    method = "gam",
    formula = y ~ s(x, k = 10),
    color = "firebrick",
    fill = "gold",
    linewidth = 0.65,
    se = TRUE
  ) +
  facet_wrap(~gene, ncol = 3, scales = "free_y") +
  labs(x = "Monocle3 pseudotime", y = "log-normalized expression") +
  theme_bw(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey90", linewidth = 0.2)
  )

pseudotime_png <- file.path(PLOT_DIR, "monocle3_marker_pseudotime_grid.png")
pseudotime_pdf <- file.path(PLOT_DIR, "monocle3_marker_pseudotime_grid.pdf")
ggsave(pseudotime_png, pseudotime_grid, width = 15, height = 12, dpi = 240)
ggsave(pseudotime_pdf, pseudotime_grid, width = 15, height = 12, device = cairo_pdf)

umap_png
umap_pdf
gene_umap_png
gene_umap_pdf
pseudotime_png
pseudotime_pdf
```
